FAT-Pointer based range addresses

### 2024-03-11 papers/ISMM/FlexPointer/paper.tex
added introduction section

@@

\section{Introduction}

\subsection{Background}

@@

\subsection{Overview}

@@

\section{Related work}
Efficient memory management, particularly in the context of 
Translation Lookaside Buffer (TLB) optimization \cite{ad-wood-2003}, has been a focal point of 
research and development within computer architecture. Various techniques have been 
proposed to mitigate TLB-related bottlenecks and improve overall system performance.

\subsection{Huge Pages}
This is used to map a very large region of memory to a 
single entry. This small/large region of memory is physically
contiguous. Most implementations of huge pages are size
aligned, For example for the x86 architecture the huge pages 
size are 4KB, 2MB and 1GB pages. 
% \subsection{Tailored page sizes}
% TODO later
% \subsection{TLB coalescing}
% This leverages the default OS allocator behavior to pack
% multiple PTEs into a single TLB entry.

\subsection{Segment}
A segment can be viewed as mapping between contiguous virtual
memory and contiguous physical memory. The property of a 
segment allows it to be larger than a page. Direct Segment
<paper reference> allows the user to set a single segment
for an application. Two registers are added to mark the start
and end of the segment. Any virtual address within this region
can be translated by adding the fixed offset between the virtual
and physical address.

\subsection{RMM}
RMM introduces the concept of adding an additional range table.
For large allocations RMM eagerly allocates contiguous physical pages.
The following allocations creates large memory ranges that are
both virtually and physically contiguous. RMM builds on the concept
of Direct segment <paper reference> by adding offset to translate a virtual address 
to physical address. RMM compares address with range boundaries 
to decide which range it belongs to. RMM queries the range table 
ofter an L1 TLB miss.

\subsection{FlexPointer}
FlexPointer is based on the RMM<paper reference RMM> paper. FlexPointer
does eagerly allocate pages which are physically contiguous and stores the ID to translate a virtual address 
to physical address on the remaining unused bits on the 64 bit virtual address. 
The paper contribution mentions shifting the TLB lookup to an earlier stage to improve
latency of accessing the TLB entries. FlexPointer immediate queries the 
range TLB for translations rather than the RMM paper which waits for the L1 TLB miss.

>>>>

% Sample citation
% \cite{ad-wood-2003}

@@

\section{Introduction}
In the dynamic landscape of computing, the pursuit of optimal performance is a constant endeavor, 
especially as applications evolve to handle increasingly complex workloads. 
One critical aspect influencing performance is memory management, where efficient 
utilization of resources is paramount. Translation Lookaside Buffers (TLBs) play a 
pivotal role in this regard, expediting memory access by storing recently accessed memory translations.
However, as applications grow in size and complexity, the capacity of TLBs often struggles to 
keep pace, leading to performance bottlenecks. To address this challenge, researchers have 
turned to innovative solutions, one of which involves harnessing the benefits of huge pages.
Huge pages, also known as large pages, allow for the allocation of memory in significantly 
larger chunks compared to traditional small pages. By reducing the number of TLB entries 
needed to access a given amount of memory, huge pages offer a potential avenue for optimizing 
TLB utilization and thereby enhancing overall system performance.

Simultaneously, advancements in hardware-level security, such as the Capability Hardware 
Enhanced RISC Instructions (CHERI) architecture, present additional opportunities for 
performance enhancement. CHERI's capability-based addressing approach not only strengthens 
system security by tightly controlling memory access but also provides avenues for 
accelerating memory management operations.

In this context, the integration of huge pages into memory management 
strategies alongside capability-based addressing in architectures like 
CHERI offers a compelling synergy. By optimizing TLB utilization through the 
utilization of huge pages and leveraging the security features of capability-based addressing, 
significant performance improvements can be realized. This approach not only enhances 
system security but also accelerates memory access

% TODO: 
% - Add references for FlexPointer, Range Memory Mapping (RMM), Direct Segment and Huge Pages
\subsection{Background}
Efficient memory management, particularly in the context of 
Translation Lookaside Buffer (TLB) optimization, has been a focal point of 
research and development within computer architecture. Various techniques have been 
proposed to mitigate TLB-related bottlenecks and improve overall system performance.

\subsubsection{Huge Pages:}
This is used to map a very large region of memory to a 
single entry. This small/large region of memory is physically
contiguous. Most implementations of huge pages are size
aligned, For example for the x86 architecture the huge pages 
size are 4KB, 2MB and 1GB pages. 
% \subsection{Tailored page sizes}
% TODO later
% \subsection{TLB coalescing}
% This leverages the default OS allocator behavior to pack
% multiple PTEs into a single TLB entry.

\subsubsection{Segment:}
A segment can be viewed as mapping between contiguous virtual
memory and contiguous physical memory. The property of a 
segment allows it to be larger than a page. Direct Segment
<paper reference> allows the user to set a single segment
for an application. Two registers are added to mark the start
and end of the segment. Any virtual address within this region
can be translated by adding the fixed offset between the virtual
and physical address.

\subsubsection{RMM:}
RMM introduces the concept of adding an additional range table.
For large allocations RMM eagerly allocates contiguous physical pages.
The following allocations creates large memory ranges that are
both virtually and physically contiguous. RMM builds on the concept
of Direct segment <paper reference> by adding offset to translate a virtual address 
to physical address. RMM compares address with range boundaries 
to decide which range it belongs to. RMM queries the range table 
ofter an L1 TLB miss.

\subsubsection{FlexPointer:}
FlexPointer is based on the RMM<paper reference RMM> paper. FlexPointer
does eagerly allocate pages which are physically contiguous and stores the ID to translate a virtual address 
to physical address on the remaining unused bits on the 64 bit virtual address. 
The paper contribution mentions shifting the TLB lookup to an earlier stage to improve
latency of accessing the TLB entries. FlexPointer immediate queries the 
range TLB for translations rather than the RMM paper which waits for the L1 TLB miss.

@@

% TODO: 
% - Add references for FlexPointer, Range Memory Mapping (RMM) and Direct segment
\subsection{Overview}
FAT-Pointers, combined with the capabilities of the CHERI (Capability Hardware Enhanced RISC Instructions) 
architecture, introduce robust memory safety and security features by incorporating additional metadata 
with memory pointers. This enhanced architecture utilizes concepts such as FlexPointer, 
Range Memory Mapping (RMM), and Direct segment to manage memory effectively.

Range addresses play a pivotal role within this framework, defining memory 
regions bounded by a starting address (BASE) and an ending address (LIMIT). 
These range addresses are encoded within FAT-pointers, allowing for precise 
control over memory access permissions.

The functionality of ranges encompasses several key aspects:
\begin{itemize}
\item \textbf{Creation of Physically Contiguous Memory Ranges}:
By defining memory regions that are physically contiguous, systems can 
achieve optimal memory access patterns, enhancing performance and efficiency.
\item \textbf{Encoding Ranges as Bounds to the Pointer}:
Integrating range bounds directly into FAT-pointers enables the architecture 
to enforce memory access restrictions at the pointer level thus allowing 
tracking of memory ranges on a pointer level.
\item \textbf{Instrumenting Block-Based Allocators with Physically Contiguous Memory}:
The integration of range-based memory concepts into memory allocation systems, such as block-based 
allocators, facilitates the efficient management and utilization of physically contiguous memory blocks, 
mitigating issues related to memory fragmentation.
\end{itemize}

@@

\section{Related work}
